Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-19 numerical-radiation-fld.tex
Reynolds edits.

In addition to the ray-tracing approach for radiation transport
described in Section~\ref{sec.num.raytracing}, \enzo\ currently includes
a field-based radiation transport solver for problems posed on uniform
(i.e.~non-AMR grids).  Detailed explanations of the model and
solution approach may be found in
\cite{NBHBROW2007}, \cite{ReynoldsHayesPaschosNorman2009} and
\cite{NRS2009}, the salient features of which are reproduced here. 
In addition, comparisons of this solver with other astrophysical
radiation transport solvers may be found in \cite{IlievEtAl2009}.

@@

We then evolve the heating and chemistry system, Equations~(\ref{eq:fld_heating})
and (\ref{eq:species_evolution}):  
$(E_r^{n+1}, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1},
{\mathrm n}_i^{n+1})$.  Due to the lack of spatial derivatives (since
advection is handled elsewhere), this system is a coupled system of
nonlinear ordinary differential equations.  This utilizes an implicit
quasi-steady-state approximation, formulated as follows.  We consider
the modified equations,
\begin{eqnarray}
  \label{eq:fld_heating_qss}
  \frac{\partial e_c}{\partial t} &=& -\frac{2\dot{a}}{a} e_c +
    \Gamma\left(\bar{E}_r,\bar{\mathrm n}_i\right) - 
    \Lambda\left(\bar{E}_r,\bar{\mathrm n}_i\right), \\
  \label{eq:fld_chemistry_qss}
  \frac{\partial {\mathrm n}_i}{\partial t} &=& k_{i,j}\left(\bar{e}_c\right)
    {\mathrm n}_e \bar{\mathrm n}_j - {\mathrm n}_i 
    \Gamma_i^{ph}\left(\bar{E}_r\right), \quad i=1,\ldots,N_s,
\end{eqnarray}
where we have defined the time-centered ``background'' states
$\bar{E}_r = \left(E_r^{n}+E_r^{n+1}\right)/2$, 
$\bar{\mathrm n}_i = \left({\mathrm n}_i^{n}+{\mathrm n}_i^{n+1}\right)/2$
and $\bar{e}_c = \left(e_c^{n}+e_c^{n+1}\right)/2$.  These equations
may each be solved {\em analytically} for their solution at the time
$t$, which we denote by
\begin{eqnarray}
  \label{eq:fld_heating_qss_sol}
  e_c(t) &=& \text{sol}_{e}\left(t,\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right), \\
  \label{eq:fld_chemistry_qss_sol}
  {\mathrm n}_i(t) &=& \text{sol}_{\mathrm n_i}
  \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right), \quad i=1,\ldots,N_s. 
\end{eqnarray}
We then define a nonlinear system of equations to compute the
time-evolved solutions $\left(E_r^{n+1}, e_c^{n+1}, 
{\mathrm n}_i^{n+1})\right)$ as
\begin{eqnarray}
  \label{eq:fld_heating_qss_fe}
  f_e(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& e_c^{n+1} -
    \text{sol}_{e}\left(t^{n+1},\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right)
    = 0, \\
  \label{eq:fld_chemistry_qss_fn}
  f_{\mathrm n_i}(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& 
    {\mathrm n}_i(t) - \text{sol}_{\mathrm n_i}
    \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right)=0, \quad
    i=1,\ldots,N_s.  
\end{eqnarray}
This system of $N_s+1$ nonlinear equations is solved using a damped
fixed point iteration, 
\[
   U_i = U_i - \lambda f_i(U), \quad i=1,\ldots,N_s+1,
\]
where $U$ is a vector containing the solutions to equations
(\ref{eq:fld_heating_qss_fe}-\ref{eq:fld_chemistry_qss_fn}).  In this
iteration, for the first 50 sweeps we use $\lambda=1$, and for more
challenging problems if that does not suffice we switch to a damping
parameter of $\lambda=0.1$.

>>>>

In addition to the ray-tracing approach for radiation transport
described in Section~\ref{sec.num.raytracing}, \enzo\ currently includes
a field-based radiation transport solver for problems posed on uniform
(i.e.~non-AMR) grids, which has been tuned for large-scale simulations
involving many ionizing sources.  Detailed explanations of the model and
solution approach may be found in
\cite{NBHBROW2007}, \cite{ReynoldsHayesPaschosNorman2009} and
\cite{NRS2009}, the salient features of which are reproduced here. 
In addition, comparisons of this solver with other astrophysical
radiation transport solvers may be found in \cite{IlievEtAl2009}.

@@

We then evolve the heating and chemistry system, Equations~(\ref{eq:fld_heating})
and (\ref{eq:species_evolution}):  
$(E_r^{n+1}, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1},
{\mathrm n}_i^{n+1})$.  Due to the lack of spatial derivatives (since
advection is handled elsewhere), this system is a coupled system of
nonlinear ordinary differential equations.  This utilizes an implicit
quasi-steady-state approximation, formulated as follows.  We consider
the modified equations,
\begin{eqnarray}
  \label{eq:fld_heating_qss}
  \frac{\partial e_c}{\partial t} &=& -\frac{2\dot{a}}{a} e_c +
    \Gamma\left(\bar{E}_r,\bar{\mathrm n}_i\right) - 
    \Lambda\left(\bar{E}_r,\bar{\mathrm n}_i\right), \\
  \label{eq:fld_chemistry_qss}
  \frac{\partial {\mathrm n}_i}{\partial t} &=& k_{i,j}\left(\bar{e}_c\right)
    {\mathrm n}_e \bar{\mathrm n}_j - {\mathrm n}_i 
    \Gamma_i^{ph}\left(\bar{E}_r\right), \quad i=1,\ldots,N_s,
\end{eqnarray}
where we have defined the time-centered ``background'' states
$\bar{E}_r = \left(E_r^{n}+E_r^{n+1}\right)/2$, 
$\bar{\mathrm n}_i = \left({\mathrm n}_i^{n}+{\mathrm n}_i^{n+1}\right)/2$
and $\bar{e}_c = \left(e_c^{n}+e_c^{n+1}\right)/2$.  These equations
may each be solved analytically for their solution at the time
$t$, which we denote by
\begin{eqnarray}
  \label{eq:fld_heating_qss_sol}
  e_c(t) &=& \text{sol}_{e}\left(t,\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right), \\
  \label{eq:fld_chemistry_qss_sol}
  {\mathrm n}_i(t) &=& \text{sol}_{\mathrm n_i}
  \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right), \quad i=1,\ldots,N_s. 
\end{eqnarray}
We then define a nonlinear system of equations to compute the
time-evolved solutions $\left(E_r^{n+1}, e_c^{n+1}, 
{\mathrm n}_i^{n+1})\right)$ as
\begin{eqnarray}
  \label{eq:fld_heating_qss_fe}
  f_e(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& e_c^{n+1} -
    \text{sol}_{e}\left(t^{n+1},\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right)
    = 0, \\
  \label{eq:fld_chemistry_qss_fn}
  f_{\mathrm n_i}(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& 
    {\mathrm n}_i(t) - \text{sol}_{\mathrm n_i}
    \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right)=0, \quad
    i=1,\ldots,N_s.  
\end{eqnarray}
This system of $N_s+1$ nonlinear equations is solved using a damped
fixed point iteration, 
\[
   U_i = U_i - \lambda f_i(U), \quad i=1,\ldots,N_s+1,
\]
where $U$ is a vector containing the solutions to equations
(\ref{eq:fld_heating_qss_fe}-\ref{eq:fld_chemistry_qss_fn}).  In this
iteration, for the first 50 sweeps we use $\lambda=1$, and for more
challenging problems if that does not suffice we switch to a damping
parameter of $\lambda=0.1$.

### 2013-05-19 numerical-radiation-homogeneous.tex
Adding AMR scaling section with an associated table.  I still need to add some references in this section.  Fixing some undefined and multiple labels and making the UVB table wider and spaced nicely.
Reynolds edits.

Homogeneous radiation field types 5 and 6 start with a spectral shape
that is then integrated against the appropriate H, He, and He$^+$
cross-sections to compute the ionization and photo-ioniziation heating
rates (we use 400 bins logarithmically spaced from 0.74 eV to $7.24
\times 10^9$ eV).  In particular, Type 5 has a hard, featureless
quasar-like power law spectrum $f_{\nu} = f_{\rm HI} \nu^{\alpha_0}$,
where $\alpha_0 = 1.5$ by default and the spectrum is normalized at
the HI ionization edge.  Type 6 has the same spectrum by is attenuated
by a column density of $10^{22}$ cm$^{-3}$ neutral hydrogen.

@@

\begin{table}
\begin{center}
\caption{Homogeneous radiation field coefficients}
\begin{tabular}{lrrrrrr}
\tableline\tableline
{Element} & {$k_0$} &  {$\alpha$} & {$\beta$} & {$\gamma$} & {$z_0$} & {$z_1$}  \\
\tableline
\multicolumn{7}{c}{Radiation Type 1 \citep{1996ApJ...461...20H} for the case $\alpha_q = 1.5$} \\
\tableline
H ionization & $6.7 \times 10^{-13}$ & 0.43 & 1/1.95 & 0 & 2.3 & 0 \\
He ionization & $6.3 \times 10^{-15}$ & 0.51 & 1/2.35 & 0 & 2.3 & 0 \\
He$^+$ ionization & $3.2 \times 10^{-13}$ & 0.50 & 1/2.00 & 0 & 2.3 & 0 \\
H heating & $4.7 \times 10^{-24}$ & 0.43 & 1/1.95 & 0 & 2.3 & 0 \\
He heating & $8.2 \times 10^{-24}$ & 0.50 & 1/2.00 & 0 & 2.3 & 0 \\
He$^+$ heating & $1.6 \times 10^{-25}$ & 0.51 & 1/2.35 & 0 & 2.3 & 0 \\
\tableline
\multicolumn{7}{c}{Radiation Type 2 \citep{1996ApJ...461...20H} for the case $\alpha_q = 1.8$} \\
\tableline
H ionization & $5.6 \times 10^{-13}$ & 0.43 & 1/1.95 & 0 & 2.3 & 0 \\
He ionization & $3.2 \times 10^{-15}$ & 0.30 & 1/2.60 & 0 & 2.3 & 0 \\
He$^+$ ionization & $4.8 \times 10^{-13}$ & 0.43 & 1/1.95 & 0 & 2.3 & 0 \\
H heating & $3.9 \times 10^{-24}$ & 0.43 & 1/1.95 & 0 & 2.3 & 0 \\
He heating & $6.4 \times 10^{-24}$ & 0.43 & 1/2.10 & 0 & 2.3 & 0 \\
He$^+$ heating & $8.7 \times 10^{-26}$ & 0.30 & 1/2.70 & 0 & 2.3 & 0 \\
\tableline
\multicolumn{7}{c}{Radiation Type 3 modified \citep{2012ApJ...746..125H}} \\
\tableline
H ionization &        $1.04 \times 10^{-12}$ & 0.231 & -0.6818 & 0.1646 & 1.855 & 0.3097 \\
He ionization &       $1.84 \times 10^{-14}$ & -1.038 & -1.1640 & 0.1940 & 1.973 & -0.6561 \\
He$^+$ ionization & $5.79 \times 10^{-13}$ & 0.278 & -0.8260 & 0.1730 & 1.973 & 0.2880 \\
H heating &            $8.86 \times 10^{-25}$ & -0.0290 & -0.7055 & 0.1884 & 2.003 & 0.2888 \\
He heating &         $5.86 \times 10^{-24}$ & 0.1764 & -0.8029 & 0.1732 & 2.088 & 0.1362 \\
He$^+$ heating &   $2.17 \times 10^{-25}$ & -0.2196 & -1.070 & 0.2124 & 1.1782 & -0.9213 \\
\end{tabular}
%\tablecomments{}
\label{table:homo_coefs}
\end{center}
\end{table}

>>>>

Homogeneous radiation field types 5 and 6 start with a spectral shape
that is then integrated against the appropriate H, He, and He$^+$
cross-sections to compute the ionization and photo-ioniziation heating
rates (we use 400 bins logarithmically spaced from 0.74 eV to $7.24
\times 10^9$ eV).  In particular, Type 5 has a hard, featureless
quasar-like power law spectrum $f_{\nu} = f_{\rm HI} \nu^{\alpha_0}$,
where $\alpha_0 = 1.5$ by default and the spectrum is normalized at
the HI ionization edge.  Type 6 has the same spectrum, but attenuated
by a column density of $10^{22}$ cm$^{-3}$ neutral hydrogen.

@@

\red{Note: we talk about several radiation field types that are not in
Table~\ref{table:homo_coefs} (4...14).  Do we want to increase the
length of the table, remove the text from the section, or perhaps just
say something in this section about how the table is for illustrative purposes? }

\begin{table*}
\begin{center}
\caption{Homogeneous radiation field coefficients}
\begin{tabular*}{0.9\textwidth}{@{\extracolsep{\fill}}lrrrrrr}
\tableline\tableline
{Element} & {$k_0$} &  {$\alpha$} & {$\beta$} & {$\gamma$} & {$z_0$} & {$z_1$}  \\
\tableline
\multicolumn{7}{c}{Radiation Type 1 \citep{1996ApJ...461...20H} for the case $\alpha_q = 1.5$} \\
\tableline
H ionization & $6.7 \times 10^{-13}$ & 0.43 & 1/1.95 & 0 & 2.3 & 0 \\
He ionization & $6.3 \times 10^{-15}$ & 0.51 & 1/2.35 & 0 & 2.3 & 0 \\
He$^+$ ionization & $3.2 \times 10^{-13}$ & 0.50 & 1/2.00 & 0 & 2.3 & 0 \\
H heating & $4.7 \times 10^{-24}$ & 0.43 & 1/1.95 & 0 & 2.3 & 0 \\
He heating & $8.2 \times 10^{-24}$ & 0.50 & 1/2.00 & 0 & 2.3 & 0 \\
He$^+$ heating & $1.6 \times 10^{-25}$ & 0.51 & 1/2.35 & 0 & 2.3 & 0 \\
\tableline
\multicolumn{7}{c}{Radiation Type 2 \citep{1996ApJ...461...20H} for the case $\alpha_q = 1.8$} \\
\tableline
H ionization & $5.6 \times 10^{-13}$ & 0.43 & 1/1.95 & 0 & 2.3 & 0 \\
He ionization & $3.2 \times 10^{-15}$ & 0.30 & 1/2.60 & 0 & 2.3 & 0 \\
He$^+$ ionization & $4.8 \times 10^{-13}$ & 0.43 & 1/1.95 & 0 & 2.3 & 0 \\
H heating & $3.9 \times 10^{-24}$ & 0.43 & 1/1.95 & 0 & 2.3 & 0 \\
He heating & $6.4 \times 10^{-24}$ & 0.43 & 1/2.10 & 0 & 2.3 & 0 \\
He$^+$ heating & $8.7 \times 10^{-26}$ & 0.30 & 1/2.70 & 0 & 2.3 & 0 \\
\tableline
\multicolumn{7}{c}{Radiation Type 3 modified \citep{2012ApJ...746..125H}} \\
\tableline
H ionization &        $1.04 \times 10^{-12}$ & 0.231 & -0.6818 & 0.1646 & 1.855 & 0.3097 \\
He ionization &       $1.84 \times 10^{-14}$ & -1.038 & -1.1640 & 0.1940 & 1.973 & -0.6561 \\
He$^+$ ionization & $5.79 \times 10^{-13}$ & 0.278 & -0.8260 & 0.1730 & 1.973 & 0.2880 \\
H heating &            $8.86 \times 10^{-25}$ & -0.0290 & -0.7055 & 0.1884 & 2.003 & 0.2888 \\
He heating &         $5.86 \times 10^{-24}$ & 0.1764 & -0.8029 & 0.1732 & 2.088 & 0.1362 \\
He$^+$ heating &   $2.17 \times 10^{-25}$ & -0.2196 & -1.070 & 0.2124 & 1.1782 & -0.9213 \\
\end{tabular*}
%\tablecomments{}
\label{table:homo_coefs}
\end{center}
\end{table*}

### 2013-05-19 numerical-radiation-raytracing.tex
Reynolds edits.

where $P$ is the photon number flux along the ray.  To sample the
radiation field at large radii, ray tracing requires at least $N_{ray}
= 4\pi R^2 / (\Delta x)^2$ rays to sample each cell with one ray,
where $R$ is the radius from the source to the cell and $\Delta x$ is
the cell width.  If one were to trace $N_{ray}$ rays out to $R$, the
cells at a smaller radius $r$ would be sampled by, on average,
$(r/R)^2$ rays, which is computationally wasteful because only a few
rays per cel  are required to provide an accurate calculation
of the radiation field (as we will show later).

>>>>

where $P$ is the photon number flux along the ray.  To sample the
radiation field at large radii, ray tracing requires at least $N_{ray}
= 4\pi R^2 / (\Delta x)^2$ rays to sample each cell with one ray,
where $R$ is the radius from the source to the cell and $\Delta x$ is
the cell width.  If one were to trace $N_{ray}$ rays out to $R$, the
cells at a smaller radius $r$ would be sampled by, on average,
$(r/R)^2$ rays, which is computationally wasteful because only a few
rays per cell  are required to provide an accurate calculation
of the radiation field (as we will show later).

### 2013-05-19 numerical-starformation.tex
Reynolds edits.

\red{( (BWO) I am a little worried that this section is too
  long. The editorial hatchet may be called for, but I think somebody
  else (who didn't write the section in the first place) needs to
  think it through. )}

@@

The Kravtsov method of star particle creation is designed to reproduce
the global Schmidt law of star formation~\citep{2003ApJ...590L...1K,
  1959ApJ...129..243S}.  This algorithm is deliberately minimal, and
is explicitly geared towards modeling star formation in a
phenomenological way on kiloparsec scales.  Stars are assume to form
with a characteristic gas timescale $\tau_*$ such that $\dot{\rho}_* =
\rho_{\rm gas}/\tau_*$ where $\rho_{\rm gas}$ and $\rho_*$ are the
baryon gas and stellar densities, respectively.  This ``constant
efficiency'' model on the scale of star formation regions is well
motivated observationally
\citep{1996AJ....112.1903Y,2002ApJ...569..157W}.  Star formation is
only allowed to take place in very dense regions with $\rho_{\rm gas}
\geq \rho_{\rm SF}$, where $\rho_{\rm SF}$ is a constant proper (as
opposed to comoving) density threshold.  No other criteria are
imposed.  Typical choices for $\tau_*$ and $\rho_{\rm SF}$ are $\tau_*
= 4$ Gyr and $\rho_{\rm SF} = 1.64~$M$_\odot$~pc$^{-3}$~$(n_{\rm H}
\sim 50$~cm$^{-3})$.  The adopted timescale is derived from the
observationally-determined normalization of the Schmidt law, and the
density threshold is determined by observations of star forming
regions on $\sim 100$ pc scales.

@@

Each ``star particle'' is assumed to represent an ensemble of stars
and is treated as a single-age stellar population (as in the previous
section).  Kravtsov assumes that the IMF is described by a Miller \&
Scalo functional form with stellar masses between $0.1$ and
$100~M_\odot$ \citep{1979ApJS...41..513M}.  All stars in this IMF with
$M_* > 8~M_\odot$ deposit $2 \times 10^{51}$ erg of thermal energy
and a mass $f_z M_*$ of metals into the cell in which they form
without delay, with $f_z \equiv \min(0.2, 0.01~M_*-0.06)$
(i.e. instantaneous deposition of metals).  The definition of $f_z$ is
a rough approximation of the results of \citet{1995ApJS..101..181W}.

@@

Feedback from Population III stars created using this algorithm comes in multiple forms.
Radiative feedback using the Moray radiation transport algorithm
\citep{Wise11_Moray} is available, using the mass-dependent hydrogen
ionizing and Lyman-Werner photon luminosities and lifetimes of the
Population III stars from \citet{2002A&A...382...28S}.  At the end of
their main-sequence lifetimes, explosion energies, ejected gas mass,
and ejected metal are caclulated using a variety of sources that
depend on the mass of the star, and are described in detail in Section
3.2.1 of \citet{2012MNRAS.427..311W}.

>>>>

% \red{( (BWO) I am a little worried that this section is too
%   long. The editorial hatchet may be called for, but I think somebody
%   else (who didn't write the section in the first place) needs to
%   think it through. )}

@@

The Kravtsov method of star particle creation is designed to reproduce
the global Schmidt law of star formation~\citep{2003ApJ...590L...1K,
  1959ApJ...129..243S}.  This algorithm is deliberately minimal, and
is explicitly geared towards modeling star formation in a
phenomenological way on kiloparsec scales.  Stars are assumed to form
with a characteristic gas timescale $\tau_*$ such that $\dot{\rho}_* =
\rho_{\rm gas}/\tau_*$ where $\rho_{\rm gas}$ and $\rho_*$ are the
baryon gas and stellar densities, respectively.  This ``constant
efficiency'' model on the scale of star formation regions is well
motivated observationally
\citep{1996AJ....112.1903Y,2002ApJ...569..157W}.  Star formation is
only allowed to take place in very dense regions with $\rho_{\rm gas}
\geq \rho_{\rm SF}$, where $\rho_{\rm SF}$ is a constant proper (as
opposed to comoving) density threshold.  No other criteria are
imposed.  Typical choices for $\tau_*$ and $\rho_{\rm SF}$ are $\tau_*
= 4$ Gyr and $\rho_{\rm SF} = 1.64~$M$_\odot$~pc$^{-3}$~$(n_{\rm H}
\sim 50$~cm$^{-3})$.  The adopted timescale is derived from the
observationally-determined normalization of the Schmidt law, and the
density threshold is determined by observations of star forming
regions on $\sim 100$ pc scales.

@@

Each ``star particle'' is assumed to represent an ensemble of stars
and is treated as a single-age stellar population (as in the previous
section).  Kravtsov assumes that the stellar initial mass function (IMF) is described by a Miller \&
Scalo functional form with stellar masses between $0.1$ and
$100~M_\odot$ \citep{1979ApJS...41..513M}.  All stars in this IMF with
$M_* > 8~M_\odot$ deposit $2 \times 10^{51}$ erg of thermal energy
and a mass $f_z M_*$ of metals into the cell in which they form
without delay, with $f_z \equiv \min(0.2, 0.01~M_*-0.06)$
(i.e. instantaneous deposition of metals).  The definition of $f_z$ is
a rough approximation of the results of \citet{1995ApJS..101..181W}.

@@

Feedback from Population III stars created using this algorithm comes in multiple forms.
Radiative feedback using the Moray radiation transport algorithm
\citep{Wise11_Moray} is available, using the mass-dependent hydrogen
ionizing and Lyman-Werner photon luminosities and lifetimes of the
Population III stars from \citet{2002A&A...382...28S}.  At the end of
their main-sequence lifetimes, explosion energies, ejected gas mass,
and ejected metal are calculated using a variety of sources that
depend on the mass of the star, and are described in detail in Section
3.2.1 of \citet{2012MNRAS.427..311W}.
